\input{ThesisPreamble}
\begin{document}

\begin{titlepage}
\begin{flushleft}
\hrule
\vspace{1 cm}
{\huge{\bf{Model Order Reduction Techniques for Hyperbolic Partial Differential Equations}}}


\vspace*{2cm}\vspace{3 cm}
{\large Chaitanya K. Shettigara}

\vspace{0.5 cm}{\today}
\end{flushleft}
\end{titlepage}

\newpage \pagenumbering{roman}
\tableofcontents

\newpage \pagenumbering{arabic}

\chapter{Introduction}
The point of this is to find more efficient ways to find the solutions to partial differential equations. 

When gravitational waves were first detected in 2015, it marked the culmination of decades of dedicated scientific and engineering effort.  It was a marvel of modern science. One of the extraordinary leaps that had been made was the use of Bayesian inverse problems for identifying the parameters of the binary black hole system that had created the gravitational wave that was detected. To do this, thousands of systems had been simulated by computational mathematicians and physicists solving the Einstein field equations. These are a set of non-linear, second order coupled partial differential equations and are extremely difficult to solve. However, to solve the inverse problem, they needed to be solved over and over again at great computational expense. In order to continue to detect gravitational waves and work out where they came from, more solutions must be found than is computationally feasible.

Model order reduction is a method that may solve this problem. The idea.


\begin{itemize}
\item The detection of gravitational waves in 2015 was a scientific triumph
\item In brought together a range of scientific and engineering disciplines all performing at the edge of human capacity. 
\item In particular, the field of inverse problems and computational mathematics. 
\item Partial differential equations describe lots of natural phenomena. 
\item Can be expensive to solve, yet sometimes this is necessary to determine outcomes for a parameter space.
\item For example, determining the likelihood function for a Bayesian parameter estimation problem (brief description of what this is)
\item Model Order Reduction seeks to reduce the computational expense
\item Reduced basis methods are one type of MOR
\item Literature mostly focusses on elliptic problems
\item Here we'll focus on hyperbolic problems
\item First scalar- scalar wave equation eg radiation patterns
\item Then vector- eg global wind patterns
\item Finally tensor- gravitational waves
\item Summarise results
\end{itemize}

\input{WaveEquation.tex}

\chapter{Spectral Methods for Solving Hyperbolic PDEs}
\begin{itemize}
\item spectral methods are advantageous in systems with large temporal or spatial variation as they are global and incorporate all the information in the domain (compared to finite difference methods)
\item using Chebyshev polynomials and Spherical Harmonics
\begin{itemize}
\item why Chebyshev- nodes placement
\item basis for
\end{itemize}
\item collocation
\item stability, accuracy and efficiency results for spectral methods for hyperbolic equations
\item illustrative examples of 1D and 3D wave equation (3D using Spherical Harmonics)
\item 
\end{itemize}



\chapter{Other Methods?}
\begin{itemize}
\item other MOR methods, I don't know what they are
\end{itemize}

\chapter{Vector valued PDEs}
\begin{itemize}
\item Hyperbolic PDEs with vector-valued solutions (2 or 3 dimensional). Maybe something aerodynamics related- global wind patterns?
\item using spin-weighted spherical harmonics
\end{itemize}

\chapter{Model Order Reduction for Vector-valued PDEs}
\begin{itemize}
\item repeat the RB and other MOR work for vector valued pdes where high fidelity solution is performed using spectral methods with spin weighted spherical harmonics.
\item accuracy, stability and efficiency results.
\end{itemize}

\chapter{2-Tensor valued PDEs}
\begin{itemize}
\item numerical relativity with 2-tensor output (metric tensor)
\item spin 2 weighted spherical harmonics
\item results from GW catalogues
\end{itemize}

\chapter{Model Order Reduction for 2-Tensor valued PDEs}
\begin{itemize}
\item repeat RB and MOR work again for NR results
\item using GW catalogues
\end{itemize}

\chapter{Conclusion}
\begin{itemize}
\item summarise results
\item suggest where results could be used.
\end{itemize}

%\appendix


%\bibliographystyle{abbrv}
%\bibliography{research.bib}

\end{document} 